---
title: "Layla Chirar 03/02/26"
subtitle: "Cleaning and Wrangling Data"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  warning: false
  message: false
---

# Introduction

This data is for commercial catching of salmon, which is important to have in order to source a sufficient amount of fish for the consumer market.

# Data

This uses a data set from Alaska Department of Fish & Game containing commercial catch data from 1878-1997 which details different salmon species. It was downloaded on 3/3/26

# Analysis

```{r}
library(dplyr)
library(tidyr)
library(readr)
```

```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

#This line opens the dataset directly from the website using the link

mean_region <- catch_original %>% #Sets up the pipe to filter by region
  select(-All, -notesRegCode) %>% #Selects all regions
  mutate(Chinook = if_else(Chinook == "I", "1", Chinook)) %>% #Changes the mistaken value in Chinook from I to 1
  mutate(Chinook = as.numeric(Chinook)) %>% #convert values to numeric
  pivot_longer(-c(Region, Year), #Make format long
               names_to = "species", #changes column name to species
               values_to = "catch") %>% #changes values to catch
  mutate(catch = catch*1000) %>% #makes total catch, not thousands
  group_by(Region) %>% #groups by region
  summarize(mean_catch = mean(catch)) %>% #calculates mean catch by year
  arrange(desc(mean_catch)) #ordered from high to low mean catch

head(mean_region) #shows the beginning of dataset
```

# Discussion

The output data doesn't reflect species, year, or the catches in the thousands. It simplifies the data into each region's mean catch per year. It shows that the SSE region had the highest number of catches and the following regions in descending order. 
