---
title: "a04"
author: "Chirar, Layla"
editor: visual
---

#tidy-data-joins assignment

```{r}
library(dplyr)
library(tidyr)
library(readr)
```

```{r}
bird_observations <- read_csv(here::here("data/raw/52_pp52_birds_1.csv"))
```

```{r}
surveys <- read_csv(here::here("data/raw/52_pp52_surveys_1.csv"))
sites <- read_csv(here::here("data/raw/52_pp52_sites_1.csv"))
taxalist <- read_csv(here::here("data/raw/52_pp52_taxalist_1.csv")) 
```

```{r}
dim(bird_observations)
head(bird_observations)
```

```{r}
glimpse(bird_observations)
summary(bird_observations)
```

```{r}
names(bird_observations)
```

```{r}
unique(bird_observations$species_id)
```

```{r}
birds_subset <- bird_observations %>%
  filter(species_id %in% c("BHCO", "RWBL")) %>%
  filter(site_id %in% c("LI-W", "NU-C"))
```

```{r}
# syntax using pipe
birds_left <- birds_subset %>%
  left_join(y = sites, by = "site_id")
# dont see x = birds_subset here because piping in birds_subset means it automatically assumes birds_subset as x.

#syntax without pipe
birds_left <- left_join(x = birds_subset, y = sites, by = "site_id")

head(birds_left)
```

```{r}
birds_full <- full_join(x = birds_subset, y=sites, by = "site_id")
head(birds_full)
```

```{r}
birds_inner <- inner_join( x = birds_subset, y = taxalist, by = 'species_id')
```

```{r}
birds_inner_left <- left_join( x = birds_subset, y = taxalist, by = 'species_id')
```

```{r}
bird_observations <- bird_observations %>%
  rename(birds_obs_notes = notes)
```

```{r}
names(bird_observations)
```

```{r}
bird_obs_subset <- bird_observations %>%
  full_join( y = taxalist, by = "species_id") %>%
  full_join( y = sites, by = "site_id") %>%
  full_join(y = surveys, by = c("site_id", "survey_id")) %>%  
  select(survey_date, common_name, park_name, bird_count, observer)

glimpse(survey_date)
```

```{r}
unique(bird_obs_subset$observer)
```

```{r}
bird_obs_subset <- bird_obs_subset %>%
  mutate(observer = if_else(condition = observer == "Josh Burns",
                            true = "J.Burns",
                            false = observer))

unique(bird_obs_subset$observer)


```

#a04

##Introduction
This data shows the bird species observed in different locations by month.

##Data
This data was collected from the central Arizona-Phoenix metropolitan area by Arizona State University researchers and was downloaded 3/9/2026.

##Analysis

```{r}
lubridate::month(surveys$survey_date)
```

```{r}
bird_obs_subset <- bird_obs_subset %>%
  mutate(survey_month = factor(lubridate::month(survey_date))) %>%
  dplyr::relocate(survey_month, .after = survey_date)
```

```{r}
bird_obs_subset <- bird_obs_subset %>%
  filter(park_name %in% c('Lindo', 'Orme', 'Palomino', 'Sonrisa'))
```

```{r}
bird_obs_park <- bird_obs_subset %>%
  group_by(park_name, survey_month) %>%
  summarize(total_birds = sum(bird_count, na.rm = TRUE), .groups = "drop")
bird_obs_park
```



##Discussion

This data shows the amount of birds observed during each month for each park. There are some rows where the survey_month is listed as N/A, which maybe indicates that there were no observations for some months.
